Refinement for open automata

### 2021-04-21 2021-Refinement/report.tex
Introducing some examples, removing pdf, correcting some line ends

\usepackage{xunicode}
\usepackage{fontspec}
\usepackage[hmargin=0.5in,marginparwidth=1.5in,includemp]{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{unicode-math}
\usepackage{tikz}
\usetikzlibrary{external}
%\usepackage{hyperref}
% \usepackage{xcolor}
\AtBeginDocument{\renewcommand\setminus{\smallsetminus}}
\usepackage{polyglossia}
\setmainlanguage{english}

@@

% possible examples for refinement: + <= |, ; <= |
% transitions between sections
% Re-read what refinement are used for

% Give the examples

@@

\section{Open Automaton}
To define an open automaton we need some preliminary definitions.
\begin{defi}[Expression algebra, Formulae, Terms]
An expression algebra \(E\) is a disjoint union \(E \defobject \terms \uplus \formulae\) where \(\terms\) is a term algebra and \(\formulae\) the formulae over \(\terms\).
\end{defi}
The terms algebra is arbitrary.
The formulae form a first order logic over \(\terms\) and some relations.
An example of term algebra can be Peano integers (zero, variable and successor function), the formulae associated can use equality relation, \(\mathit{sum}\mpar{a, b, c} \defobject a = b + c\) relation and \(\mathit{prod}\mpar{a, b, c} \defobject a = b \times c\) relation.

\begin{defi}[Free variables, Closed terms, Closed formulae]
\defitem \(\fvars{e}\) is the set of unbound variables in \(e \in E\).
\defitem The expressions restricted to variables in \(V\) are \(E_V \defnotation \mset{e \in E}{\fvars{e} = \emptyset}\); \(E_V \subset E\).
\defitem The closed expressions are expressions restricted to variables in \(\emptyset\), \(E_\emptyset\).
\end{defi}
By restriction, terms and formulae restricted to variables and closed versions are also defined.
A closed expression is an expression without unbound variable.
We can use the previous example of first order formula on Peano integers to illustrate these definitions.
\(x\), \(S\mpar{y}\) are valid terms, \(0\) is a closed term, \(\fvars{x} = \mbrc{x}, \fvars{S\mpar{y}} = \mbrc{y}, \fvars{0} = \emptyset\).
\(x = 0\), \(\forall y, \exists x, \mathit{sum}\mpar{y, z, x}\) are valid formulae, \(\forall x, \neg S\mpar{x} = 0\) is a valid closed formula, \(\fvars{x = 0} = \mbrc{x}, \fvars{\forall y, \exists x, \mathit{sum}\mpar{y, z, x}} = \mbrc{z}, \fvars{\forall x, \neg S\mpar{x} = 0} = \emptyset\).
\begin{defi}[Interpretation, Values, Satisfiability, (Parallel) substitution]
We assume that the following are given:
\defitem The values \(\values\), which are interpretations of closed terms.
\defitem The validity relation on closed formulae, \({\vdash} \subseteq \cformulae\).
\defitem The substitution in \(e \in E\) of \(x \in \fvars{e}\) by \(t \in \terms\), \(e\subst{t}{x}\).
\defitem The parallel substitution in \(e \in E\) of variables in \(V\) by \(\psi: V \to \terms\), \(e\psubst{\psi}\).
\end{defi}
For the parallel substitution, the set \(V\) is not required to be a subset of \(\fvars{e}\).
In the case where it is not, the variables in \(V \setminus \fvars{e}\) are not substituted.
The substitutions might give a nonsensical expression; for instance let the terms be integers and pairs with (pointwise) addition, \(\mpar{a + b}\psubst{a \mapsto 7, b \mapsto \mpar{4, 5}}\) is a ill-formed term.
This can be guarded with the atom \(e\subst{t}{x} \in E\) and \(e\psubst{\psi} \in E\) and it will implicitely be the case, for instance when there is quantification on \(t\) and \(\psi\), to simplify notations.

The interpretation of terms is supposed to be decidable.
The validity of formulae might not be decidable nor complete nor consistent, however we will pretend like they are because these are really hard problems for logicians that we don't want to deal with.\footnote{In practise a term algera with only equality is used, and it is has all these properties.}

A value \(v\) may be used for keeping a variable state, and then injected in terms for substitution.
While this is correct when \(\values \subseteq \cterms\), in the other cases this will be used as a shorthand for substitution with any term which can be interpreted as \(v\).
We suppose that the interpretation of terms is compatible w.r.t.\@ substitution, that is if two terms \(t\), \(t'\) are interpreted with the same value, then replacing \(t\) by \(t'\) in a well-formed expression makes an equivalent well-formed expression.
A quantifier followed by a finite set will be used as a shorthand for the quantification on every variable in the set, for instance \(\forall A, \exists B, P\) with \(A = \mbrc{a_1, \dots, a_\card{A}}, B = \mbrc{b_1, \dots, b_\card{B}}\) means \(\forall a_1, \dots, \forall a_\card{A}, \exists b_1, \dots, \exists b_\card{B}, P\).

The validity relations will be noted using the classical relational notation \(\vdash f \defnotation {f \in {\vdash}}\).
The usual validity of a formula \(f \in \formulae\) is noted \({\vdash f} \defnotation {\vdash \forall \fvars{f}, f}\).
The usual validity of a formula \(f \in \formulae\) with some fixed valuation \(\sigma: V \to \values\) is noted \({\sigma \vdash f} \defnotation {\vdash \forall\mpar{\fvars{f} \setminus V}, f\psubst{\sigma}}\).

From there the expression algebra is fixed.
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
A semantic open automaton is a tuple \(\OA{S}{i}{V}{\varphi}{J}{T}\) with \(S\) the set of states, \(i \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\varphi: V \to \values\) the initial valuation of variables, \(J\) the set of hole names and \(T\) a set of open-transitions.

\(S, V, J\) are arbitrary sets, only \(J\) is required to be finite.
\end{defi}
The variable names may clash when considering two automata, in that case we suppose that we can still distinguish the variables in the formulas.
All the possible open transitions, from which transitions in \(T\) are selected, are defined below.
\begin{defi}[Open transition]
A semantic open transition is a tuple \nmm{\OT{s}{s'}{\alpha\mpar{v_\alpha}}{\mpar{\beta_j\mpar{v_{\beta j}}}^{j \in J'}}{g}{\psi}} with \(s, s' \in S\) the source and target states, \(\alpha \in \actions\) an action label, \(J' \subseteq J\) the holes involved, \(\beta_j \in \actions\) the action labels of the holes, \(v_\alpha \in \rterms{V}\) the emitted action, \(v_{\beta j}^{j \in J'} \in \rterms{V}\) the hole actions, \(g \in \rformulae{V}\) the guard and \(\psi: V \to \rterms{V}\) the new value of the variables.
\end{defi}
The set of action labels \(\actions\) is arbitrary and implicit.
To simplify formulae, the notations \(\alpha \defnotation \alpha\mpar{v_\alpha}\) and \(\alpha = \beta \defnotation \alpha = \beta \wedge v_\alpha = v_\beta\) will be used.

To illustrate these definitions let's look at two implementations (figure \ref{fig:enable}) of the LOTOS\marginpar{cite something here} operator enable in the open automata model.
The enable operator runs its the left hand side agent until it chooses to finish, at which point it produces an action \(\delta\) that is synchronised with the first action of the right hand side agent which must be \(\act{acc}\), then only the latter agent runs.
During the synchronised action \Quentin{data can be exchanged}{I need to read a little bit more on that; is that a return code?}. % TODO: read
\begin{figure}
\centering
\input{enable_state.tex}
\vrule
\input{enable_var.tex}
\caption{Enable operator, on the left state oriented, on the right variable oriented.}
\label{fig:enable}
\end{figure}
\begin{exi}[Enable, state-oriented]
To draw automata, the usual circle for states and simple arrows for transitions convention is used.
The initial state is indicated by a double circle.
States names are indicated inside the circles and transitions labels are drawn near their corresponding arrow.

The open transitions do not indicate the source and target states since that is the role of the transitions arrows; only the action is on the bottom side of the open transition.
Also in the example, transitions are specified using variables in place of actions or values.
This is used to represent all the transitions created by a correct substitution of these variables, as there are often infinitly many.

There is a model where open automata have finitely many transitions and variables like that are allowed in the expressions and in the action data, not as a tool to graphically represent an infinity of transitions but as a real construct.
This model is the model of structural open automata as opposed to the model used here of semantic open automata.
The structural version is more adapted to uses where the object manipulated must be finite, for example in algorithms, whereas the semantic version avoids the manipulation of substitutions, which simplifies proofs.

Let's get back to the example.
The open automata drawn on the left is \(\OA{\mbrc{L, R}}{L}{\emptyset}{\mbrc{}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are all the valid transitions for any substitution of \(x\) and \(y\) of
\begin{align*}
	\OT{L}{L}{x}{\mbrc{l \mapsto x}}{x \neq \delta}{\mbrc{}} &&
	\OT{L}{R}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{x = y}{\mbrc{}} &&
	\OT{R}{R}{x}{\mbrc{r \mapsto x}}{\top}{\mbrc{}}
\end{align*}
% TODO: Explain example
\end{exi}
\begin{exi}[Enable, variable-oriented]
% TODO: Explain example
\end{exi}

Now we can define some utilitary functions:
\begin{defi}[Guard, Out-transition]
\(\fOT{s}\) are called the out-transitions of \(s\).
\begin{align*}
	\fguard{\OTg} & \defnotation g &
	\fOT{s} & \defnotation \mbrc{\OTg \in T}
\end{align*}
\end{defi}
The intuition of a semantic open automaton is a partially defined LTS with variables, guards on transitions and parametrised actions.
From this point semantic open automata and open transitions will be called automata and transitions for simplicity.

\section{Semantic and composition of Open Automata} % This is taken from 2007.10770.pdf
% TODO: Intro
When the auomaton is in a state \(s \in S\) with a valuation of its variables \(\sigma: V \to \values\), it cannot perform transitions \nmm{\OTg \in T} such that \(\neg g\psubst{\sigma}\).
After performing the transition and emitting action \(\alpha\), the automaton is in the state \(s'\) with valuation \(\psi\psubst{\sigma}\).
The holes are synchronised using a handshake like mecanism when the transition happens.
\begin{exi}
\end{exi}

An open automaton can be partially specified, the partial specification comes from the holes.
A hole can be filled with an open automaton, this operation is called \Quentin{composition}{Or substitution, or filling, what do you prefer?}.
\begin{defi}[Composition of open automata]
The composition of \(A_c \defobject \OA{S_c}{i_c}{V_c}{\varphi_c}{J_c}{T_c}\) in the hole \(k \in J_p\) of \(A_p \defobject \OA{S_p}{i_p}{V_p}{\varphi_p}{J_p}{T_p}\) is
\begin{align*}
	A_p\subst{A_c}{k} \defnotation & \OA{S_p \times S_c}{\mpar{i_p, i_c}}{V_p \uplus V_c}{\varphi_p \uplus \varphi_c}{J_c \uplus J_p \setminus \mbrc{k}}{T} \\
	\text{With } T \defobject & \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p \setminus \mbrc{k}}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{\OTx{p}{}{}{p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p}, k \notin J', s_c \in S_c}
\end{align*}
Similarly to expression substitution, the parallel composition is noted \(A\psubst{A_j^{j \in J}}\).
\end{defi}
The actions emitted when \(A_c\) makes a transition is sychronised with the action of the hole \(k\) in transitions of \(A_p\) which have it as a hole actions (first transition set, \(\alpha_c = \beta_k\)).
No transition of \(A_c\) is performed when a transition that do not refer to the hole \(k\) is performed in \(A_p\) (second transition set, \(k \notin J'\)).
The composition may look like a handshake, with both automata running in parallel (product of states and joint variables) but it is actually not symmetric.
\begin{exi}
\end{exi}

@@

The relations will have to be at least preorders, simulations and forbid the introduction of deadlocks (state where no transition is possible) to be called refinement simulation.
On top of that weak refinement are refinement where some actions, typically one named \(\tau\), are considered not observable and do not have to be matched.
In the weak refinement the relations have to forbid the introduction of livelocks (only a sequence of \(\tau\) possible) and respect \(\tau\)-stuttering \marginpar{cite many things arround here}.
While being preorder is simple to express for a relation, being a simulation, do not introduce new deadlocks/livelocks and \(\tau\)-stuttering are properties about the states of the related automata, therefore they are more complex to express.
We will need a preliminary concept before defining them.
\begin{defi}[Relation under predicate]
A relation under predicate between states of two automata \(\OA{S_1}{i_1}{J_1}{V_1}{\varphi_1}{T_1}\) and \(\OA{S_2}{i_2}{J_2}{V_2}{\varphi_2}{T_2}\) is a function in \(S_1 \times S_2 \to \rformulae{V_1 \uplus V_2}\).
Two states \(s_1 \in S_1, s_2 \in S_2\) are related under a valuation \(\sigma: V_1 \uplus V_2 \to \values\) iff \(\sigma \vdash R\mpar{s_1, s_2}\).
\end{defi}
Now let's adapt the properties about states of automata to open automata.
\begin{defi}[Simulation on open automata]
A relation \(\leq\) is a simulation if for any related automata \(\wrel{A_1}{A_2}{R}\) where \(A_1 \defobject \OA{S_1}{i_1}{J_1}{V_1}{\varphi_1}{T_1}\), \(A_2 \defobject \OA{S_2}{i_2}{J_2}{V_2}{\varphi_2}{T_2}\) and \(R: S_1 \times S_2 \to \rformulae{V_1 \uplus V_2}\), \(R\) satisfies both
\defitem Initial states are related under initial valuations: \(\varphi_1 \uplus \varphi_2 \vdash R\mpar{i_1, i_2}\);
\defitem From related states, all out-transitions from \(A_1\) can be simulated in \(A_2\) and their target states are related:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \sigma: V_1 \uplus V_2 \to \values, \\
	\bigsymb{\forall} \OTx{1}{}{1}{1} \in T_1, \mpar{\sigma \vdash R\mpar{s_1, s_2} \wedge g_1} \implies \bigsymb{\exists} \OTx{2}{}{2}{2} \in T_2, \\
	\sigma \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
The last formula means that for every pair of related states under valuation of the automaton variables and every possible transitions (\(\sigma \vdash g_1\)) from the first automaton, there is a possible transition (\(\sigma \vdash g_2\)) such that the produced action matches (\(\alpha_1 = \alpha_2\)), the holes actions from same hole names (\(J'_1 \cap J'_2\)) match (\(\beta_{1j} = \beta_{2j}\)) and the target states are related after variable update.
This is a natural extenion of the notion of simulation on LTS, which is the same definition without variables and holes and guards.
It will serve as a watchdog for any relation that will be defined with ``simulation" in its name, except weak-simulation where the transition from the first automaton that produce \(\tau\) actions are excluded.

>>>>

\usepackage{xunicode}
\usepackage{fontspec}
\usepackage[hmargin=0.5in,marginparwidth=1.5in,includemp]{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{unicode-math}
\usepackage{tikz}
\usetikzlibrary{external}
% \usepackage{hyperref}
% \usepackage{xcolor}
\AtBeginDocument{\renewcommand\setminus{\smallsetminus}}
\usepackage{polyglossia}
\setmainlanguage{english}

@@

% possible examples for refinement: + <= |, ; <= |
% transitions between sections
% Re-read what refinement are used for, LOTOS
% read CCSL and BIP

@@

\section{Open Automaton}
To define an open automaton we need some preliminary definitions.
\begin{defi}[Expression algebra, Action algebra, Formulae, Terms]
An expression algebra \(E\) is a disjoint union \(E \defobject \terms \uplus \actions \uplus \formulae\) of the terms, the actions and the formulae.

The terms \(\terms\) is a term algebra.
The action algebra \(\actions\) is the set of pairs \(\labels \times \terms\) where \(\labels\) are the action labels.
The formulae \(\formulae\) are at least the first order formulae over \(\terms\) and \(\labels\).
\end{defi}
The term algebra and action labels are arbitrary.
An element \(\mpar{l, t} \in \actions\) is noted \(l\mpar{t}\).
The formulae form at least a first order logic with an equality relation that is not necessarily the equality on \(\terms\).

An example of term algebra can be Peano integers (zero, variable and successor function), the formulae associated can use equality relation, \(\mathit{sum}\mpar{a, b, c} \defobject a = b + c\) relation and \(\mathit{prod}\mpar{a, b, c} \defobject a = b \times c\) relation.

\begin{defi}[Free variables, Closed terms, Closed formulae]
\defitem \(\fvars{e}\) is the set of unbound variables in \(e \in E\).
\defitem The expressions restricted to variables in \(V\) are \(E_V \defnotation \mset{e \in E}{\fvars{e} \subseteq V}\); \(E_V \subset E\).
\defitem The closed expressions are expressions restricted to variables in \(\emptyset\), \(E_\emptyset\).
\end{defi}
Terms, actions and formulae restricted to variables and their closed versions are also defined by restriction.
A closed expression can contain variables, only unbound variable are restricted.

We can use the previous example of first order formula on Peano integers to illustrate these definitions.
\(x\), \(S\mpar{y}\) are valid terms, \(0\) is a closed term, \(\fvars{x} = \mbrc{x}\), \(\fvars{S\mpar{y}} = \mbrc{y}\), \(\fvars{0} = \emptyset\).
\(x = 0\), \(\forall y, \exists x, \mathit{sum}\mpar{y, z, x}\) are valid formulae, \(\forall x, \neg S\mpar{x} = 0\) is a valid closed formula, \(\fvars{x = 0} = \mbrc{x}\), \(\fvars{\forall y, \exists x, \mathit{sum}\mpar{y, z, x}} = \mbrc{z}\), \(\fvars{\forall x, \neg S\mpar{x} = 0} = \emptyset\).

\begin{defi}[Interpretation, Values, Satisfiability, (Parallel) substitution]
We assume that the following are given:
\defitem The values \(\values\), which are interpretations of closed terms.
\defitem The validity relation on closed formulae, \({\vdash} \subseteq \rformulae\).
\defitem The substitution in \(e \in E\) of \(x \in \fvars{e}\) by \(t \in \terms\), \(e\subst{t}{x}\).
\defitem The parallel substitution in \(e \in E\) of variables in \(V\) by \(\psi: V \to \terms\), \(e\psubst{\psi}\).
\end{defi}
For the parallel substitution, the set \(V\) is not required to be a subset of \(\fvars{e}\).
In the case where it is not, the variables in \(V \setminus \fvars{e}\) are not substituted.
The substitutions might give a ill-formed expression; for instance let the terms be integers and pairs with (pointwise) addition, \(\mpar{a + b}\psubst{a \mapsto 7, b \mapsto \mpar{4, 5}}\) is a ill-formed term.
This can be guarded with the atom \(e\subst{t}{x} \in E\) and \(e\psubst{\psi} \in E\) and it will implicitely be the case, for instance when there is quantification on \(t\) and \(\psi\), to simplify notations.

The interpretation of terms is supposed to be decidable.
The validity of formulae might not be decidable nor complete nor consistent, however we will pretend like they are because these are really hard problems for logicians that we don't want to deal with.
In practise the formulae will be given to a SMT solver and we make sure they have all the previous properties.

A value \(v\) may be used for keeping a variable state, and then injected in terms for substitution.
While this is correct when \(\values \subseteq \rterms\), in the other cases this will be used as a shorthand for substitution with any term which can be interpreted as \(v\).
We suppose that the interpretation of terms is compatible w.r.t.\@ substitution, that is if two terms \(t\), \(t'\) are interpreted with the same value, then replacing \(t\) by \(t'\) in a well-formed expression makes an equivalent well-formed expression.
In a formula, a quantifier followed by a finite set will be used as a shorthand for the quantification on every variable in the set, for instance \(\forall A, \exists B, P\) with \(A = \mbrc{a_1, \dots, a_\card{A}}, B = \mbrc{b_1, \dots, b_\card{B}}\) means \(\forall a_1, \dots, \forall a_\card{A}, \exists b_1, \dots, \exists b_\card{B}, P\).
% maybe Valuation as formula /\ var=val

The validity relations will be noted using the classical relational notation \(\vdash f\).
The precedence of \(\vdash\) is the lowest on the right side and higher than \(\uplus\) on the left side: \(\forall a \, b, a \uplus b \vdash x \wedge y \implies \exists z, P\mpar{x,z}\) is the same as \(\forall a \, b, \mpar{\mpar{a \uplus b} \vdash \mpar{\mpar{x \wedge y} \implies \exists z, P\mpar{x,z}}}\).
The usual validity of a formula \(f \in \formulae\) is noted \({\vdash f} \defnotation {\vdash \forall \fvars{f}, f}\).
The usual validity of a formula \(f \in \formulae\) with some fixed valuation \(\sigma: V \to \values\) is noted \({\sigma \vdash f} \defnotation {\vdash \forall \fvars{f\psubst{\sigma}}, f\psubst{\sigma}}\). % TODO: remettre en exists? faire des LHS plus compliques style forall LHS exists RHS
% maybe satisfiability with formulas on the LHS (and variabl sets)

From there the expression algebra is fixed.
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
A semantic open automaton is a tuple \(\OA{S}{s_0}{V}{\sigma_0}{J}{T}\) with \(S\) the set of states, \(s_0 \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\sigma_0: V \to \values\) the initial valuation of variables, \(J\) the set of hole names and \(T\) a set of open-transitions.

\(S, V, J\) are arbitrary finite sets.
\end{defi}
The variable names may clash when considering two automata, in that case we suppose that we can still distinguish the variables in the formulas.
All the possible open transitions, from which transitions in \(T\) are selected, are defined below.
\begin{defi}[Open transition]
A semantic open transition is a tuple \nmm{\OTg} with \(s, s' \in S\) the source and target states, \(\alpha \in \ractions\) the produced action, \(J' \subseteq J\) the holes involved, \(\beta_j \in \ractions\) the actions of the holes, \(g \in \rformulae[V]\) the guard and \(\psi: V \to \rterms\) the new value of the variables.
\end{defi}
The intuition of a semantic open automaton is a partially defined LTS with variables, guards on transitions and parametrised actions.
Initially the automaton is in the initial state with the initial valuation.
In any state with any valuation, it can perform transitions which source state is the current state and which guard is valid in the current valuation if the holes emit the indicated action.
By performing the transition the automaton emits the indicated action and updates its variables and state.

\begin{figure}
\centering
\input{enable_state.tex}
\vrule
\input{enable_var.tex}
\caption{Enable operator implementation with open automata, on the left state oriented, on the right variable oriented}
\label{fig:enable}
\end{figure}
To illustrate these definitions let's look at two implementations (figure \ref{fig:enable}) of the LOTOS\marginpar{cite something here} operator enable in the open automata model.
The enable operator runs its the left hand side agent until it chooses to finish, at which point it produces an action \(\delta\) that is synchronised with the first action of the right hand side agent which must be \(\act{acc}\), then only the latter agent runs.
During the synchronised action \Quentin{data can be exchanged}{I need to read a little bit more on that; is that a return code?}.
\begin{exi}[Enable, state-oriented]
To draw automata, the usual circle for states and simple arrows for transitions convention is used.
The initial state is indicated by a double circle.
States names are indicated inside the circles and transitions labels are drawn near their corresponding arrow.

The open transitions do not indicate the source and target states since that is the role of the transitions arrows; only the action is on the bottom side of the open transition.
Also in the example, transitions are specified using variables in place of actions or values.
This is used to represent all the transitions created by a correct substitution of these variables, which cannot be drawn otherwise as there are often infinitly many.

There is a model where open automata have finitely many transitions and variables like that are allowed in any expression, not only as a tool to graphically represent an infinity of transitions but as a real construct.
This model is the structural open automata as opposed to the model used here of semantic open automata.
The structural version is more adapted to uses where the object manipulated must be finite, for example in algorithms, whereas the semantic version avoids the manipulation of substitutions, which simplifies proofs.

Let's get back to the example.
The automata on the left side is \(\OA{\mbrc{L, R}}{L}{\emptyset}{\mbrc{}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are generated by all valid substitution of \(x\) and \(y\) in
\begin{align*}
	\OT{L}{L}{x}{\mbrc{l \mapsto x}}{x \neq \delta}{\mbrc{}} &&
	\OT{L}{R}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{x = y}{\mbrc{}} &&
	\OT{R}{R}{x}{\mbrc{r \mapsto x}}{\top}{\mbrc{}}
\end{align*}
Note that we could have expressed the transition in the middle with \(\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{x}}\) wich would have avoided many transitions with trivially false guards like the one where \(x \mapsto 1, y \mapsto 2\) which has the guard \(1 = 2\).

This automaton is an implementation of the enable operator because it begins in the state \(L\) where it allows any transition from its hole \(l\) which is not \(\delta\) then synchronises its holes on the same data, effectively allowing a data exchange when it goes into state \(R\), and finally allows any transition from its hole \(r\).
Only the synchronisation is hidden to the exterior by sending a \(\tau\) which is usually a silent action always allowed, unsynchronised with other holes and passed unmodified.
Well here \(\tau\) is not always allowed from the holes in the state \(L\) the hole \(r\) cannot emmit it for the sake of simplifying the first example of open automaton.
\end{exi}
\begin{exi}[Enable, variable-oriented]
The automata drawn on the right side of the figure is an alternative implementation of the enable operator which has already been proven equivalent in previous work by ...\marginpar{cite sth here}.
The drawing introduces the double arrows which is used to denote the initial valuation.
The automata drawn is \(\OA{\mbrc{\_}}{\_}{\mbrc{v}}{l^\mbrc{v}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are like previously all the transitions generated by valid substitution of \(x\) and \(y\) in
\begin{align*}
	\OT{\_}{\_}{x}{\mbrc{l \mapsto x}}{v = l \wedge x \neq \delta}{\mbrc{}} &&
	\OT{\_}{\_}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{v = l \wedge x = y}{\mbrc{v \gets r}} &&
	\OT{\_}{\_}{x}{\mbrc{r \mapsto x}}{v = r}{\mbrc{}}
\end{align*}
Note that in this example \(r\) is a hole name and also a closed term, but not a variable.
The same applies for \(l\).
\(v\) is a variable name for the automaton and a variable in the guard, but its value is not substituted.
If an automaton variable \(v\) is in a transition outside of the guard then it would mean that, any transition obtained by valid substitution of \(v\) with a value \(x\) and by adding \(v = x\) to the guard, would be a transition of the automaton.

% TODO: run the automaton conceptually
\end{exi}

Finally we can define some utilitary functions:
\begin{defi}[Guard, Out-transition]
\(\fOT{s}\) are called the out-transitions of \(s\).
\begin{align*}
	\fguard{\OTg} & \defnotation g &
	\fOT{s} & \defnotation \mbrc{\OTg \in T}
\end{align*}
\end{defi}
From this point semantic open automata and open transitions will be called automata and transitions for simplicity.

\section{Semantic and composition of Open Automata} % This is taken from 2007.10770.pdf
% TODO: Intro

An open automaton can be partially specified, the partial specification comes from the holes.
A hole can be filled with an open automaton, this operation is called \Quentin{composition}{Or substitution, or filling, what do you prefer?}.
\begin{defi}[Composition of open automata]
The composition of \(A_c \defobject \OA{S_c}{s_{0c}}{V_c}{\sigma_{0c}}{J_c}{T_c}\) in the hole \(k \in J_p\) of \(A_p \defobject \OA{S_p}{s_{0p}}{V_p}{\sigma_{0p}}{J_p}{T_p}\) is
\begin{align*}
	A_p\subst{A_c}{k} \defnotation & \OA{S_p \times S_c}{\mpar{s_{0p}, s_{0c}}}{V_p \uplus V_c}{\sigma_{0p} \uplus \sigma_{0c}}{J_c \uplus J_p \setminus \mbrc{k}}{T} \\
	\text{With } T \defobject & \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p \setminus \mbrc{k}}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{\OTx{p}{}{}{p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p}, k \notin J', s_c \in S_c}
\end{align*}
Similarly to expression substitution, the parallel composition is noted \(A\psubst{A_j^{j \in J}}\).
\end{defi}
The actions emitted when \(A_c\) makes a transition is sychronised with the action of the hole \(k\) in transitions of \(A_p\) which have it as a hole actions (first transition set, \(\alpha_c = \beta_k\)).
No transition of \(A_c\) is performed when a transition that do not refer to the hole \(k\) is performed in \(A_p\) (second transition set, \(k \notin J'\)).
The composition may look like a handshake, with both automata running in parallel (product of states and joint variables) but it is actually not symmetric.
\begin{exi}
\begin{figure}
\centering
\input{Traffic_Lights_Spec.tex}
\caption{The specification of a traffic light system}
\label{fig:tls}
\end{figure}
\begin{figure}
\centering
\input{Traffic_Lights_Controller.tex}
\vrule
\input{Traffic_Lights_Register.tex}
\caption{On the left: The controller agent; On the right: The counter agent}
\label{fig:tlh}
\end{figure}
\begin{figure}
\centering
\input{Traffic_Lights_Full.tex}
\caption{The full traffic lights system}
\label{fig:tlf}
\end{figure}
\end{exi}

@@

The relations will have to be at least preorders, simulations and forbid the introduction of deadlocks (state where no transition is possible) to be called refinement simulation.
On top of that weak refinement are refinement where some actions, typically one named \(\tau\), are considered not observable and do not have to be matched.
In the weak refinement the relations have to forbid the introduction of livelocks (only a sequence of \(\tau\) possible) and respect \(\tau\)-stuttering \marginpar{cite many things arround here}.
While being preorder is simple to express for a relation, being a simulation, do not introduce new deadlocks/livelocks and \(\tau\)-stuttering are properties about the states of the related automata, therefore they are more complex to express.
We will need a preliminary concept before defining them.
\begin{defi}[Relation under predicate]
A relation under predicate between states of two automata \(\OA{S_1}{s_{01}}{J_1}{V_1}{\sigma_{01}}{T_1}\) and \(\OA{S_2}{s_{02}}{J_2}{V_2}{\sigma_{02}}{T_2}\) is a function in \(S_1 \times S_2 \to \rformulae[V_1 \uplus V_2]\).
Two states \(s_1 \in S_1, s_2 \in S_2\) are related under a valuation \(\sigma: V_1 \uplus V_2 \to \values\) iff \(\sigma \vdash R\mpar{s_1, s_2}\).
\end{defi}
Now let's adapt the properties about states of automata to open automata.
\begin{defi}[Simulation on open automata]
A relation \(\leq\) is a simulation if for any related automata \(\wrel{A_1}{A_2}{R}\) where \(A_1 \defobject \OA{S_1}{s_{01}}{J_1}{V_1}{\sigma_{01}}{T_1}\), \(A_2 \defobject \OA{S_2}{s_{02}}{J_2}{V_2}{\sigma_{02}}{T_2}\) and \(R: S_1 \times S_2 \to \rformulae[V_1 \uplus V_2]\), \(R\) satisfies both
\defitem Initial states are related under initial valuations: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\);
\defitem From related states, all out-transitions from \(A_1\) can be simulated in \(A_2\) and their target states are related:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \sigma: V_1 \uplus V_2 \to \values, \\
	\bigsymb{\forall} \OTx{1}{}{1}{1} \in T_1, \mpar{\sigma \vdash R\mpar{s_1, s_2} \wedge g_1} \implies \bigsymb{\exists} \OTx{2}{}{2}{2} \in T_2, \\
	\sigma \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
The last formula means that for every pair of related states under valuation of the automaton variables and every possible transitions (\(\sigma \vdash g_1\)) from the first automaton, there is a possible transition (\(\sigma \vdash g_2\)) such that the produced action matches (\(\alpha_1 = \alpha_2\)), the holes actions from same hole names (\(J'_1 \cap J'_2\)) match (\(\beta_{1j} = \beta_{2j}\)) and the target states are related after variable update.
This is a natural extenion of the notion of simulation on LTS, which is the same definition without variables and holes and guards.
It will serve as a watchdog for any relation that will be defined with ``simulation" in its name, except weak-simulation where the transition from the first automaton that produce \(\tau\) actions are excluded.
